The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-16 talk/vmil2012/paper.tex
More fixes

Our aim is to help understand the constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points, meaning the recorded
and compiled path has to be left returning control to the interpreter. They need
enough associated information to enable rebuilding the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of this particular
tracing JIT and clarify the rationale behind the design decisions based on the
numbers provided here.

@@

There are several aspects to be taken into account w.r.t. guards, the first aspect is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace, – which is
the common case – should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the construction of a tracing just-in-time compiler.

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind the implementation of guards in
RPython's tracing just-in-time compiler. the contributions of this paper are:
\begin{itemize}
  \item An analysis and benchmark of guards in the context of RPython's tracing JIT.
  %An analysis of guards in the context of RPython's tracing JIT to
  %substantiate the aforementioned observation, based on a set of benchmarks,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level components of the JIT and describe the rationale behind the design.
\end{itemize}

@@

After tracing has been finished the trace is optimized.
During optimization a large percentage of operations can be removed.\todo{add a reference to the figure showing the optimization rates?}
In the process the resume data is transformed into its final, compressed form.
The rationale for not compressing the resume data during tracing
is that a lot of guards will be optimized away.
For them, the compression effort would be lost.

@@

The possible source of information are:

@@

Figure~\ref{fig:benchmarks} extends Figure~\ref{fig:guard_percent} and summarizes the total number of operations that were
recorded during tracing for each of the benchmarks and what percentage of these
operations are guards. The number of operations was counted on the unoptimized
and optimized traces. The Figure also shows the overall optimization rate for
operations, which is between 69.4\% and 83.89\%, of the traced operations and the
optimization rate of guards, which is between 65.8\% and 86.2\% of the
operations. This indicates that the optimizer can remove
most of the guards, but after the optimization pass these still account for
15.2\% to 20.2\% of the operations being compiled and later executed.
The frequency of guard operations makes it important to store the associated
information efficiently and also to make sure that guard checks are executed
quickly.

@@

\subsection{Guard Failures}
\label{sub:guard_failure}
The last point in this discussion is the frequency of guard failures.
Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
relative amounts of guards that ever fail and of guards that fail often enough that a bridge is compiled.
\footnote{
    The threshold used is 200 failures. This rather high threshold was picked experimentally to give
    good results for long-running programs.
}

@@

Linking side exits to pieces of later compiled machine code was described first
in the context of Dynamo~\cite{Bala:2000wv} under the name of fragment linking.
Once a new hot trace is emitted into the fragment cache it is linked to the side
exit that led to the compilation of the fragment. Fragment Linking avoids the
performance penalty involved in leaving the compiled code. Fragment linking
also allows to remove compensation code associated to the linked fragments that
would have been required to restored the execution state on the side exit.

>>>>

Our aim is to help understand the constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points, meaning the recorded
and compiled path has to be left returning control to the interpreter. They need
enough associated information to enable rebuilding the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, to explain how they are implemented in the different levels of this particular
tracing JIT and to clarify the rationale behind the design decisions based on the
numbers provided here.

@@

There are several aspects to be taken into account w.r.t. guards, the first of them is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace, – which is
the common case – should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the construction of a tracing just-in-time compiler.

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind the implementation of guards in
RPython's tracing just-in-time compiler. the contributions of this paper are:
\begin{itemize}
  \item an analysis and benchmark of guards in the context of RPython's tracing JIT,
  %An analysis of guards in the context of RPython's tracing JIT to
  %substantiate the aforementioned observation, based on a set of benchmarks,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level components of the JIT and describe the rationale behind the design
\end{itemize}

@@

After tracing has been finished the trace is optimized.
During optimization a large percentage of operations can be removed. \todo{add a reference to the figure showing the optimization rates?}
In the process the resume data is transformed into its final, compressed form.
The rationale for not compressing the resume data during tracing
is that a lot of guards will be optimized away.
For them, the compression effort would be lost.

@@

The possible sources of information are:

@@

Figure~\ref{fig:benchmarks} extends Figure~\ref{fig:guard_percent} and summarizes the total number of operations that were
recorded during tracing for each of the benchmarks and what percentage of these
operations are guards. The number of operations was counted on the unoptimized
and optimized traces. The figure also shows the overall optimization rate for
operations, which is between 69.4\% and 83.89\%, of the traced operations and the
optimization rate of guards, which is between 65.8\% and 86.2\% of the
operations. This indicates that the optimizer can remove
most of the guards, but after the optimization pass these still account for
15.2\% to 20.2\% of the operations being compiled and later executed.
The frequency of guard operations makes it important to store the associated
information efficiently and also to make sure that guard checks are executed
quickly.

@@

\subsection{Guard Failures}
\label{sub:guard_failure}
The last point in this discussion is the frequency of guard failures.
Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
relative amounts of guards that ever fail and of guards that fail often enough that a bridge is compiled.\footnote{
    The threshold used is 200 failures. This rather high threshold was picked experimentally to give
    good results for long-running programs.
}

@@

Linking side exits to pieces of later compiled machine code was described first
in the context of Dynamo~\cite{Bala:2000wv} under the name of fragment linking.
Once a new hot trace is emitted into the fragment cache it is linked to the side
exit that led to the compilation of the fragment. Fragment linking avoids the
performance penalty involved in leaving the compiled code. Fragment linking
also allows to remove compensation code associated to the linked fragments that
would have been required to restored the execution state on the side exit.
